### 2018-03-28 paper/sexSelectionSpacing-ver1.tex
Edits of results and conclusion sections

To overcome this problem, I condition on the likelihood of parity progression 
when examining birth spacing.
If, for example, women with a given set of characteristics have an 80\%
chance of ending a spell with a birth their median spell duration is when
their survival curve shows that 40\% are predicted to have exited the spell.%
\footnote{
It is important to note that this is not the same as simply calculating 
the median \emph{observed} spell lengths among women who have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.
}
Median durations can then be averaged across women with different 
characteristics using the probabilities of ending a spell with a child
as weights.
How well this approach works depends on whether the spell length covered is 
sufficiently long that only very few women would be likely to give
birth later.
I discuss the choice of spell length below.
In addition to calculating conditional spell length for different 
percentages, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

@@

I exclude visitors to the household, as well as
women who have been married more than once, divorced, or who are 
not living with their husband,
women with inconsistent age at marriage,
and those with missing information on education.
Women interviewed in NFHS-3 or NFHS-4 who were never married or where gauna 
had not been performed were also dropped.
The same goes for women who had at least one multiple births,
reported having a birth before age 12, had a birth before marriage, or
duration between births of less than nine months.
Women who reported less than nine months between marriage and first birth
remain in the sample---unless they are dropped for another reason---because 
10 to 20 percent fall into this category across the four surveys.
Although it is possible that some of these births are premature, the high number of
women who report a birth less than six months after their marriage indicates that for
a majority conception likely occurred before marriage.

Finally, I restrict the sample to Hindus,
who constitute about 80 percent of India's population.
If the use of sex selection differ between Hindu and other religions, such 
as Sikhs, assuming that the baseline hazard is the same would lead to bias.
The other groups are each so small relative to Hindus that it is not
possible to estimate different baseline hazards for each group.
Furthermore, the other groups are so different in background and son 
preference that combining them into one group would not make sense.

@@

As mentioned above, 10 to 20 percent of women gave birth less than nine months
after they were married, and the first spell, therefore, begins at the 
month of marriage.
The exception is for women married very young, where I use the month they 
turned 12.
The second and subsequent spells begin nine months after the previous birth 
because that is the earliest we should expect to observe a new birth.
A few women report births that occurred less than nine months 
after the previous birth; I drop those women from the sample as mentioned above.

All spells continue until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place;
sterilization of the woman or her husband;
or too few births are observed for the method to work.
For the first spell I set censoring at 72 months (six years) after marriage,
while for the second and third spells censoring is at 72 months and 
for the fourth spell 57 months after a birth can occur.
With nine month between the previous birth and the beginning of
the next spell, the maximum durations between births I consider are
6.75 years for the second and third spells and 5.5 years for the 
fourth spell.

@@

Appendix Table \ref{tab:des_stat1} presents descriptive statistics for
the spells by education level and period they began.
There is a substantial number of censored observations.
As an example, for highly educated women who had their first child in the 2005--2016
period, more than 40\% did not have their second child by the time of the survey.
Censoring becomes even more important for the third and fourth
spells, with more than 70 percent of the observations censored.
The level of censoring also increases with parity and time period,
which reflects a combination of factors: timing of the surveys
relative to the periods of interest, later beginning of childbearing, 
falling fertility, and the increase in spell lengths from 
sex-selective abortions.

@@

The population has become substantially better educated over time.
Women with no education constituted almost 60 percent 
in the first period, but less than twenty percent in the last period.
Correspondingly, in the first period just over twenty percent had eight or more 
years of education, whereas in the last period more than 60\% did.
These changes are an underestimate of the increase in female
education overall because many of the younger women with more education
have not yet married than therefore are not in the sample.%
\footnote{
Women with eight or more years of education accounted for 65.9\% of
unmarried women in NFHS-3 and 82.8\% of unmarried women in NFHS-4.
See \citet[p 56]{International-Institute-for-Population-Sciences-IIPS2007}
and \citet[p 61]{International-Institute-for-Population-Sciences-IIPS2017}
for more information.
 }

@@

Second, do the shapes of survival curves differ across sex compositions 
within a spell?
The shapes of the survival curves provides clues to whether parents are
still trying to conceive earlier after girls than boys.
I expect this to be the case if there is still a strong preference for
boys, and, as discussed above, this behavior may even become more pronounced
with access to sex selection.
At a technical level, the shapes of the survival curve serve as an
indicator for the presence of non-proportionality.

Finally, does the sex ratio change within a spell?
There are two reasons why this is interesting.
First, if the sex ratio changes within a spell simply using the observed sex
ratio at birth---as is standard practice in the literature---will not provide
correct estimates of the use of sex selection or the sex ratio when child
bearing is completed.
Second, the development in the sex ratio within a spell provides clues
to parents' preferences over the ideal spacing between births and the
distribution of son preference in the sample.

@@

The predicted median duration shown is the weighted average over the women in the 
sample's predicted median durations, using the individual probabilities of a birth 
as weights.
For each woman in a sample, I calculate her probability of ending the spell 
with a birth, and her contribution to median spacing is how long it will take
from the beginning of the spell to where she is predicted to reach half of that probability.
For example, if a woman has a 80\% probability of having a birth by the end of 
the spell, her median duration is the number of months it takes to reach
40\% probability of having exited with a birth (equivalent to the 60\% point
in a standard survival curve because that captures how many have not yet
exited).

For second through fourth spell I show whether durations for sex composition
other than only girls are statistically significantly different from the
duration with only girls.
The statistical significance is based on the bootstrapped differences in
durations across sex compositions.
For each new sample the difference in duration is calculated and these
differences are used to calculate the standard error and the level of
significance for the difference in the original sample.

The cleanest test is comparing durations after only boys with durations after
only girls, but the number of births to women with only sons becomes small 
in the later periods because of falling fertility.
Hence, it is possible to have substantial differences in spacing that are
not statistically significant because of low power, especially for the third 
and fourth spell.
It is, in principle, possible to estimate the model taking into account the 
ordering of children, but this would further lower the power of the test, by 
adding one extra group for the third spell and four extra groups for the fourth spell.

The predicted sex ratio is the weighted average over the women in the sample's 
individual predicted sex ratios, using the individual probabilities of a birth by 
the end of the spell as weights.
Each woman's predicted sex ratio is the weighted average of the predicted
percentage boys over the months in the spell calculated using equation 
(\ref{eq:probability_son}) and the probability of giving birth in each month 
as weights.%
\footnote{
Imagine a spell has two periods and that the estimated percentage boys for a
woman are 54\% and 66\% and that the likelihood of having a birth is 20\% and 40\%
(measured as the change in the survival curve; in other words, there is 
40\% chance that she does not have a birth by the end of this spell).
This woman's percentage boys is then $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$.
}
The predicted sex ratio captures the percent boys that will be born to women 
in the sample once child bearing for that spell is over.

Each predicted precent boys is tested against the natural percentage 
boys using the bootstrapped standard errors.
The natural sex ratio is approximately 105 boys to 100 girls or
51.2\% \citep{ben-porath76b,jacobsen99,Portner2015b}.
The predicted percentage boys may differ from the natural rate because of 
natural variation, any remaining recall error not corrected for, or 
sex selection.

@@

Predicted median duration has generally increased over time within
each sex composition.
The exception is for the first spell, where there is little
change, and possibly even a small decline.
This decline may be the result of a combination of lower recall
error, increasing age at marriage, and better health.

The differences in duration across sex compositions are more 
pronounced with higher-order spells.
For the second spell, the difference in median duration between
having a girl or a boy as the first-born are mostly between 0.5 and 1 
month, although it varies from a low of 0.2 months for rural women 
in 1995--2004 period to a high of 1.5 months
for urban women in the 1985--1994 period.
Focusing on rural women for the third spell, the difference is,
on average, slightly over one months between only girls and only boys 
and the differences are statistically significant in all periods.
For rural women the differences for the fourth spell are just below
2 months, with the exception of the 1972--1984 period, which is
also the only period where the difference is not statistically significant.

It is possible that remaining recall error is behind some of the
smaller differences in duration.
One example is the second spell in the 1995--2004 period, where there 
are only 0.2 months difference between duration after one girl and 
duration after one boy.
Similarly, the small difference for the fourth spell in the 1972--1984
period may be the result of recall error with a very high 
percentage boys for women with three girls, despite no availability
of sex selection.

@@

Women with one to seven years of education follow a pattern
broadly similar to those with no education as 
Table \ref{tab:median_sex_ratio_med} shows.
There is a more equal distribution in the number of women
across urban and rural for this education group than for the
no education women.
As a result, the only spell where there are few women to base 
results on are for the fourth spell for urban women and the first 
period of the fourth spell for rural women.

Spacing between births increased over the four decades.
On average, the increase is approximately two months.
The exception is again the first spell, where the time from
marriage to first birth remained the same or fell slightly.

In the period before sex selection became available, 1972--1985,
the differences in spacing between only girls and only boys
increase with spell number.
For both urban and rural women the difference in duration
is just over a month for the second spell, but for the
third spell it is 1.4 and 2.3 months for urban and rural
women, respectively.
In the three later periods there is no clear pattern
in the differences across sex compositions.

@@

The decreases in differences are accompanied by increases
in predicted sex ratios, some of which are statistically 
significantly different from the natural sex ratio.
Hence, even though the evidence is not strong, it does
appear that women with one to seven years of education 
have begun to use sex selection at higher-order spells.
This is also consistent with the falling fertility for both
urban and rural women shown below.

@@

Women with eight or more years of education appear the least likely
to follow the pattern of shorter spacing when they have only girls
compared to only boys before sex selection became available as 
Table \ref{tab:median_sex_ratio_high} shows.
Part of the problem may be that there are few women with that level
of education for the 1972--1984 period.
This means that only the results for the first and second spells in 
both urban and rural areas are likely to be reliable.
Furthermore, there are relatively few women with three boys who
go on to having a fourth birth, both in urban and rural areas.

The spacing between births increases substantially over time, especially
for urban women. 
Urban women show an increase in median duration of between a quarter
and three-quarters of a year, whereas the increase for rural women is 
around a quarter of a year.
The spacing between marriage and the first birth is still the
exception, and for this group of women there is a decline of
about a month for both urban and rural areas.

In urban areas, the increase in spacing with only girls is so 
substantial that women with only girls have longer spacing to the 
next birth than any of the other sex compositions from 1985 onward.%
\footnote{
The exception is the second spell for the 1985--1994 period, where women
with one girl has 0.4 months longer spacing the women with one boy.
}
This reversal in the standard spacing pattern when there is strong
son preference is substantial, especially for the third and fourth 
spells.
For the latest period the difference between only girls than the
other sex compositions are between 3.4 and 5.4 months for the third
and fourth spell.
Furthermore, the differences are statistically significant, except
for only boys for the fourth spell because few women have a fourth
birth if they have three boys already.

Rural women also show a reversal of the normal spacing pattern,
although the changes are less pronounced than for urban women.
The third spell show increasingly longer spacing with only
girls compare to the other two sex compositions over time with
most differences statistically significant, but the differences
are smaller than for urban women.
For the second spell there is little difference in spacing with
one girl and with one boy and there are few rural women to 
based the fourth spell results on.

The predicted sex ratios at the end of the spells show that
the reversal in spacing patterns is not the result of a declining son
preference but rather corresponds to a substantially more male-biased 
sex ratio.
For urban women the predicted sex ratio with only girls 
is consistently above 60\% boys for the third and fourth spells.
For rural women the predicted sex ratios are lower than for urban
women, but stil substantial and statistically significant at close 
to 60\%.
The high predicted sex ratios are presumably the result of increased 
use of sex selection.

When trying to understand the strength of son preference it is interesting
that the sex ratio is also statistically significantly different
from the natural rate in the case where women already have one
son for the third and fourth spells.
Again, the sex ratio in the presence of one son for the third and
fourth spells are higher in urban areas than rural areas, although
the difference is less than for women with only girls.
Hence, it is possible that women are still willing to use sex selection
even after giving birth to one. 
Alternatively, this behavior may be in response to either experienced
or expected mortality of the first son born.

It also appears that there is a slightly elevated sex ratio among 
first-born for urban women. 
There is, however, no corresponding increase in spacing; rather
the median duration has fallen over time and is the lowest among
the three education groups.
Hence, unless duration between marriage and first birth has decreased 
substantially by itself, for example, because of improved health, it
is not clear how much stock to place on the increased sex ratio
among first-borns as evidence for sex selection among first-borns.

Both of the last two results are different from prior studies using NFHS data.
There are two potential explanations.
First, NFHS-4 is substantially larger than the three prior surveys.
Hence, it is possible that the effects have been there all along, but 
we simply did not have the power to detect them.
Second, it is possible that, at least, the first-born results are an
artifact of any recall error not captured by the data restrictions.
Certain types of recall error would also explain why the sex ratio for
the second spell and one boy is statistically significantly \emph{below} 
the natural sex ratio \citep{Portner2018b}.

@@

Median spacing is a convenient way to understand the major developments,
but may hide important changes in the distribution of spacing.
This section therefore provides more detail on how spacing is distributed
over time and over sex compositions using a graphical approach.

I show survival curves conditional on parity progression rather than
standard survival curves. 
The advantage of this approach is that it is possible to directly compare 
the distribution of spacing to next birth, independently of differences in
how likely the next birth is across groups.
In other words, these results reflect the spacing we should observe once
all childbearing is over for the women in the sample.
Because the conditional survival curves are independent of the likelihood
of parity they all begin at 100\% and end at 0\%.

Instead of averaging across the entire sample, I calculate the conditional
survival curves for a representative woman using the method detailed in 
Section \ref{sec:strategy}.
The representative woman characteristics are the means of continuous explanatory 
variables and the majority category for categorical explanatory variables.
Common to all representative women is that they do not own land and belong to neither 
a scheduled caste nor a scheduled tribe.

In the interest of brevity, I discuss only a representative subset.
Figure \ref{fig:pps_spell1} shows spacing within the first spell across the
four periods for rural women with no education and for urban women with 
eight or more years of education.
Figures \ref{fig:pps_low} and \ref{fig:pps_high} show spacing across sex
compositions for the second, third, and fourth spells for the first period, 
1972--1984, and the last period, 2005--2016, for rural women with no 
education and urban women with eight or more years of education.
Standard and conditional survival curves for all groups and time periods 
are shown in the Appendix, and I discuss some examples of standard survival
curves in the following section.

@@

The duration from marriage to first birth is only affected by son preferences if
people use sex selection to achieve a son as their first-born.
As I discuss above, the biased sex ratio for the first spell for the most educated 
women does suggest that there might be sex selection on the first birth.
I therefore begin by comparing in Figure \ref{fig:pss_spell1} how the distribution 
of spacing has changed over time for rural women with no education and urban women 
with eight or more years of education.
What is most striking is how similar the distribution of spacing is across the
two groups.

Both groups show little change in median spacing, but that masks a substantial
compression of when most of the births occur.
In the 1972--1984 period the middle 80\% of births for women with no education 
are predicted to occur between approximately 6 and 52 months, whereas in the 
2005--2016 period it is between 12 and 46 months.
This is equivalent to a compression of a full year.
Women with eight or more years of education show a slightly smaller compression:
in the 1972--1984 period the middle 80\% of births are predicted to occur between
approximately 6 and 44 months, whereas in the 2005--2016 period it is between
9 and 40 months, or a compression of seven months.

The pattern for both groups shows us that women are less likely to conceive 
before marriage in the later time periods than earlier.
In the first two periods there was a relatively smooth decline in number of
women without a birth starting at the time of marriage, but in the last two
periods there is now few women exiting early after marriage and instead a 
substantial dip between 9 and 12 months after marriage.

It is likely that the compressed spacing, beginning at nine months
after marriage in the later periods, is associated with better health
and higher age at marriage.
The descriptive statistics above show, for example, that women without 
education has seen an increase in the average age at marriage from below 
16 to 18.5 from the first period to the last while women with the most 
education increased from 19.5 to 20.4.

@@

For women with no education Figure \ref{fig:pps_low} shows that the median results
above are consistent with the distribution of spacing within a spell with spacing 
after only girls consistently shorter than after only boys.
There are slight variations in how strong this effect is across the three spells.
On the one had, for the second spell the difference in spacing between first child a girl 
versus first child a boy as first-born remains almost the same across the four
decades.
On the other hand, for the third spell the differences between two girls and 
the other sex compositions become more pronounced over time, and especially
the difference between two girls and two boys have become substantially larger.
The fourth spell results are based on relatively few observations, but do indicate
that only girls and two girls and one boy is associated with shorter spacing than
only boys or one girl and two boys.
Although the evidence is not overly strong, it does appear that there is a slight
preference for two boys in the sense that spacing becomes longer once the couple
has at least two boys than for the other sex compositions.

@@

The distribution of spacing, shown in Figure \ref{fig:pps_high}, for the second spell 
conditional survival curve for urban women with eight or more years of education 
is almost indistinguishable from that of rural women with no education 
for the 1972--1984 period.
This points to these two very different groups of women behaving in a similar fashion
in response to son preference in a situation where there was little access to
prenatal sex determination or little incentive to use it early even for those 
who might have had access.

Whereas the no education group show almost no change across the four decades,
the high education women show a rightward shift if the spacing pattern when
the first child is a girl.
The result is that the spacing patterns are close to identical after first-born 
girls and after first-born boys.
Given the elevated predicted percentage boys show in Table 
\ref{tab:median_sex_ratio_high} this change is consistent with use of 
sex selection for women with a girl as their first child.

The hypothesized correlation between use of sex selection and reversal of 
the standard spacing pattern show up clearly in the 2005--2016 period for the 
third spell with spacing to the third child consistently longer with two girls 
than with either two girls or one boy and one girl.
The rightward movement for only girls from the, mostly, standard pattern in 
the 1972--1984 period to the new pattern in the 2005--2016 period is substantial.
There is also indications that the spacing distribution has moved rightward
for women with one boy and one girl relatively to women with two boys, which
provides some evidence for use of sex selection for this group.
The caveat is that the 1972--1985 period also had an biased sex ratio, which
points to uncorrected recall error for this group.

>>>>

To overcome this problem, I condition on the likelihood of parity progression 
when examining birth spacing.
If, for example, women with a given set of characteristics have an 80\%
chance of ending a spell with a birth their median spell duration is when
their survival curve shows that 40\% are predicted to have exited the spell.%
\footnote{
It is important to note that this is not the same as simply calculating 
the median \emph{observed} spell lengths among women who have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.
}
Median durations can then be averaged across women with different 
characteristics using the probabilities of ending a spell with a child
as weights.
How well this approach works depends on whether the spell length covered is 
sufficiently long that few women would be likely to give birth later.
I discuss the choice of spell length below.
In addition to calculating conditional spell length for different 
percentages, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

@@

I exclude visitors to the household, as well as
women who have been married more than once, divorced, or who are 
not living with their husband,
women with inconsistent age at marriage,
and those with missing information on education.
Women interviewed in NFHS-3 or NFHS-4 who were never married or where gauna 
had not been performed were also dropped.
The same goes for women who had at least one multiple births,
reported having a birth before age 12, had a birth before marriage, or
duration between births of less than nine months.
Women who reported less than nine months between marriage and first birth
remain in the sample---unless they are dropped for another reason---because 
8.5\% fall into this category across the four surveys.
Although it is possible that some of these births are premature, the high number of
women who report a birth less than six months after their marriage indicates that for
a majority conception likely occurred before marriage.

Finally, I restrict the sample to Hindus,
who constitute about 80\% of India's population.
If the use of sex selection differ between Hindu and other religions, such 
as Sikhs, assuming that the baseline hazard is the same would lead to bias.
The other groups are each so small relative to Hindus that it is not
possible to estimate different baseline hazards for each group.
Furthermore, the other groups are so different in background and son 
preference that combining them into one group would not make sense.

@@

As mentioned above, 8.5\% of women gave birth less than nine months
after they were married, and the first spell, therefore, begins at the 
month of marriage.
The exception is for women married very young, where I use the month they 
turned 12.
The second and subsequent spells begin nine months after the previous birth 
because that is the earliest we should expect to observe a new birth.
A few women report births that occurred less than nine months 
after the previous birth; I drop those women from the sample as mentioned above.

All spells continue until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place;
sterilization of the woman or her husband;
or too few births are observed for the method to work.
For the first spell I set censoring at 108 months (nine years) after marriage,
while for the second, third, and fourth spells censoring is at 96 months 
(eight years) after a birth can occur.
TK number of births observed after these periods.
With these cut-offs, less than one percent of observed occur after
the spell cut-off.

@@

Appendix Table \ref{tab:des_stat1} presents descriptive statistics for
the spells by education level and period they began.
There is a substantial number of censored observations.
As an example, for highly educated women who had their first child in the 2005--2016
period, more than 40\% did not have their second child by the time of the survey.
Censoring becomes even more important for the third and fourth
spells, with more than 70\% of the observations censored.
The level of censoring also increases with parity and time period,
which reflects a combination of factors: timing of the surveys
relative to the periods of interest, later beginning of childbearing, 
falling fertility, and the increase in spell lengths from 
sex-selective abortions.

@@

The population has become substantially better educated over time.
Women with no education constituted almost 60\% 
in the first period, but less than twenty percent in the last period.
Correspondingly, in the first period just over twenty percent had eight or more 
years of education, whereas in the last period more than 60\% did.
These changes are an underestimate of the increase in female
education overall because many of the younger women with more education
have not yet married than therefore are not in the sample.%
\footnote{
Women with eight or more years of education accounted for 65.9\% of
unmarried women in NFHS-3 and 82.8\% of unmarried women in NFHS-4.
See \citet[p 56]{International-Institute-for-Population-Sciences-IIPS2007}
and \citet[p 61]{International-Institute-for-Population-Sciences-IIPS2017}
for more information.
 }

@@

Second, do the shapes of survival curves differ across sex compositions 
within a spell?
The shapes of the survival curves provide clues to whether parents are
still trying to conceive earlier after girls than boys.
I expect this to be the case if there is still a strong preference for
boys, and, as discussed above, this behavior may even become more pronounced
with access to sex selection.
At a technical level, the shapes of the survival curve serve as an
indicator for the presence of non-proportionality.

Finally, does the sex ratio change within a spell?
There are two reasons why this is interesting.
First, if the sex ratio changes within a spell, relying on the observed sex
ratio at birth---as is standard practice in the literature---will not provide
correct estimates of the use of sex selection or the sex ratio when 
childbearing is completed.
Second, any changes in the sex ratio within a spell provides clues
to parents' preferences over the ideal spacing between births and the
distribution of son preference in the sample.

@@

The predicted median duration shown is the weighted average over the women in the 
sample's predicted median durations, using the individual probabilities of 
giving birth as weights.
For each woman in a sample, I calculate her probability of ending the spell 
with a birth, and her contribution to median spacing is how long it will take
from the beginning of the spell to where she is predicted to reach half of that probability.
For example, if a woman has an 80\% probability of having a birth by the end of 
the spell, her median duration is the number of months it takes to reach
40\% probability of having exited with a birth.
The 40\% is equivalent to the 60\% point in a standard survival curve because that 
captures how many have not yet exited.

For second through fourth spell I show whether durations for sex composition
other than only girls are statistically significantly different from the
duration with only girls.
I use bootstrapping to calculate the level of statistical significance.
For each new sample, I calculate the difference in duration and these
differences are used to calculate the standard error and the level of
significance for the difference in the original sample.

The cleanest test is comparing durations after only boys with durations after
only girls, but the number of births to women with only sons becomes small 
in the later periods because of falling fertility.
Hence, it is possible to have substantial differences in spacing that are
not statistically significant because of low power, especially for the third 
and fourth spell.
It is, in principle, possible to estimate the model taking into account the 
ordering of children, but this would further lower the power of the test, by 
adding one additional group for the third spell and four additional groups for the fourth spell.

The predicted sex ratio is the weighted average over the women in the sample's 
individual predicted sex ratios, using the individual probabilities of having
given birth by the end of the spell as weights.
Each woman's predicted sex ratio is the weighted average of the predicted
percentage boys over the months in the spell calculated using equation 
(\ref{eq:probability_son}) and the probability of giving birth in each month 
as weights.%
\footnote{
Imagine a spell has two periods and that the estimated percentage boys for a
woman are 54\% and 66\% and that the likelihood of having a birth is 20\% and 40\%.
The likelihood of having a birth is the change in the survival curve; 
in other words, there is 40\% chance that she will not have given birth by the end of 
this spell.
This woman's percentage boys is then $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$.
}
The predicted sex ratio captures the percent boys that will be born to women 
in the sample once childbearing for that spell is over.

Each predicted percent boys is tested against the natural percentage 
boys using the bootstrapped standard errors.
The natural sex ratio is approximately 105 boys to 100 girls or
51.2\% \citep{ben-porath76b,jacobsen99,Portner2015b}.
The predicted percentage boys may differ from the natural rate because of 
natural variation, any remaining recall error not corrected for, or 
sex selection.

@@

Predicted median duration has increased over time within
each sex composition.
The exception is for the first spell, where there is little
change, and possibly even a small decline.
This decline may be the result of a combination of lower recall
error, increasing age at marriage, and better health.

The differences in duration across sex compositions increase with 
spell number.
For the second spell, the difference in median duration between
having a girl or a boy as the first-born are mostly between 0.5 and 1 
month, although it varies from a low of 0.2 months for rural women 
in 1995--2004 period to a high of 1.5 months
for urban women in the 1985--1994 period.
Focusing on rural women for the third spell, the difference is in
duration, on average, slightly over one month between only girls and only boys 
and the differences are statistically significant in all periods.
For rural women, the differences for the fourth spell are just below
two months, except for the 1972--1984 period, which is
also the only period where the difference is not statistically significant.

It is possible that remaining recall error is behind some of the
smaller differences in duration.
One example is the second spell in the 1995--2004 period, where there 
is only 0.2 months difference between duration after one girl and 
duration after one boy.
Similarly, the small difference for the fourth spell in the 1972--1984
period may be the result of recall error with a very high 
percentage of boys for women with three girls, despite no availability
of sex selection.

@@

Women with one to seven years of education follow a pattern
broadly similar to those with no education as 
Table \ref{tab:median_sex_ratio_med} shows.
The distribution across urban and rural is more even for this education
group than for the no education women.
As a result, the only spells where there are few women to base 
results on are for the fourth spell for urban women and the first 
period of the fourth spell for rural women.

The spacing between births increased over the four decades.
On average, the increase is approximately two months.
The exception is again the first spell, where the time from
marriage to first birth remained the same or fell slightly.

In the period before sex selection became available, 1972--1985,
the differences in spacing between only girls and only boys
increase with spell number.
For both urban and rural women, the difference in duration is just over
a month for the second spell, but it is 1.4 and 2.3 months for urban and
rural women, respectively, for the third spell.
In the three later periods, there is no clear pattern
in the differences across sex compositions.

@@

The decreases in differences are accompanied by increases
in predicted sex ratios, some of which are statistically 
significantly different from the natural sex ratio.
Hence, even though the evidence is not substantial, it does
appear that women with one to seven years of education 
have begun to use sex selection at higher-order spells.
An emerging use of sex selection is also consistent with the falling
fertility for both urban and rural women shown below.

@@

Women with eight or more years of education appear the least likely
to follow the pattern of shorter spacing when they have only girls
compared to only boys before sex selection became available as 
Table \ref{tab:median_sex_ratio_high} shows.
Part of the problem may be that there are few women with that level
of education for the 1972--1984 period.
Hence, only the results for the first and second spells in 
both urban and rural areas are likely to be reliable.
Furthermore, there are relatively few women with three boys who
go on to having a fourth birth, both in urban and rural areas.

The spacing between births increases substantially over time, especially
for urban women. 
Urban women show an increase in median duration of between a quarter
and three-quarters of a year, whereas the increase for rural women is 
around a quarter of a year.
The spacing between marriage and the first birth is still the
exception, and for this group of women, there is a decline of
about a month for both urban and rural areas.

In urban areas, the increase in spacing with only girls is so 
substantial that women with only girls have longer spacing to the 
next birth than any of the other sex compositions from 1985 onward.%
\footnote{
The exception is the second spell for the 1985--1994 period, where women
with one girl have 0.4 months longer spacing the women with one boy.
}
This reversal in the standard spacing pattern is substantial, 
especially for the third and fourth spells.
For the latest period, the difference between only girls than the
other sex compositions is between 3.4 and 5.4 months for the third
and fourth spell.
Furthermore, the differences are statistically significant, except
for only boys for the fourth spell because few women have a fourth
birth if they have three boys already.

Rural women also show a reversal of the typical spacing pattern,
although the changes are less pronounced than for urban women.
The third spell shows that spacing for women with only girls become
increasing longer compared to the other two sex compositions.
Most of the differences are statistically significant, although they are 
smaller than for urban women.
For the second spell, there is little difference in spacing for
women with one girl and women with one boy.

The predicted sex ratios at the end of the spells show that
the reversal in spacing patterns is not the result of a declining son
preference but instead corresponds to a substantially more male-biased 
sex ratio.
For urban women, the predicted sex ratio with only girls 
is consistently above 60\% boys for the third and fourth spells.
For rural women, the predicted sex ratios are lower than for urban
women but still substantial and statistically significant at close 
to 60\%.
The high predicted sex ratios are presumably the result of increased 
use of sex selection.

When trying to understand the strength of son preference, it is interesting
that the sex ratio is also statistically significantly different
from the natural rate in the case where women already have one
son for the third and fourth spells.
Again, the sex ratio in the presence of one son for the third and
fourth spells are higher in urban areas than rural areas, although
the difference is less than for women with only girls.
Hence, it is possible that women are still willing to use sex selection
even after giving birth to one. 
Alternatively, this behavior may be in response to either experienced
or expected mortality of the first son born.

It also appears that there is a slightly elevated sex ratio among 
first-born for urban women. 
There is, however, no corresponding increase in spacing; instead,
the median duration has fallen over time and is the lowest among
the three education groups.
Hence, unless duration between marriage and first birth has decreased 
substantially by itself, for example, because of improved health, it
is not clear how much stock to place on the increased sex ratio
among first-borns as evidence for sex selection among first-borns.

Both of the last two results are different from prior studies using NFHS data.
There are two potential explanations.
First, the NFHS-4 is substantially larger than the three prior surveys.
Hence, it is possible that the effects have been there all along, but 
we did not have the power to detect them.
Second, it is possible that, at least, the first-born results are an
artifact of any recall error not captured by the data restrictions.
Certain types of recall error would also explain why the sex ratio for
the second spell and one boy is statistically significantly \emph{below} 
the natural sex ratio \citep{Portner2018b}.

@@

Median spacing is a convenient way to understand the overall changes but
may hide important differences in the distribution of spacing.
This section, therefore, provides more detail on how spacing is distributed
over time and sex compositions using a graphical approach.

I show survival curves conditional on parity progression rather than
standard survival curves. 
The advantage of this approach is that it is possible to directly compare 
the distribution of spacing to next birth across groups, independently of 
differences in how likely the next birth is.
It does, however, depend on covering a long enough spell period
to ensure that only relatively few births are likely to take place after the
censoring date as discussed above.
Because the conditional survival curves are independent of the likelihood
of final parity progression, they all begin at 100\% and end at 0\%.

Instead of averaging across the entire sample, I calculate the conditional
survival curves for an average woman using the method detailed in 
Section \ref{sec:strategy}.
The average woman characteristics are the means of continuous explanatory 
variables and the majority category for categorical explanatory variables.
Common to all is that they do not own land and belong to neither 
a scheduled caste nor a scheduled tribe.

In the interest of brevity, I discuss only a representative subset.
Figure \ref{fig:pps_spell1} shows spacing for the first spell across the
four periods for rural women with no education and urban women with 
eight or more years of education.
Figures \ref{fig:pps_low} and \ref{fig:pps_high} show spacing across sex
compositions for the second, third, and fourth spells for the first period, 
1972--1984, and the last period, 2005--2016, for rural women with no 
education and urban women with eight or more years of education.
The Appendix shows standard and conditional survival curves for all groups and 
time periods, and I discuss some examples of standard survival
curves in the following section.

@@

The duration from marriage to first birth is only affected by son preferences if
people use sex selection to achieve a son as their first-born.
As I discuss above, the biased sex ratio for the first spell for the most educated 
women does suggest that there might be sex selection on the first birth.
I, therefore, begin by comparing in Figure \ref{fig:pss_spell1} how the distribution 
of spacing has changed over time for rural women with no education and urban women 
with eight or more years of education.
What is most striking is how similar the distribution of spacing is across the
two groups.

Both groups show little change in median spacing, but that masks a substantial
compression of when most of the births occur.
In the 1972--1984 period, the middle 80\% of births for women with no education 
are predicted to occur between approximately 6 and 52 months, whereas in the 
2005--2016 period it is between 12 and 46 months.
Hence, the compression is equivalent to a full year.
Women with eight or more years of education show a slightly smaller compression:
in the 1972--1984 period, the middle 80\% of births are predicted to occur between
approximately 6 and 44 months, whereas in the 2005--2016 period it is between
9 and 40 months, or a compression of seven months.

The pattern for both groups shows us that women are less likely to conceive 
before marriage in the later time periods than earlier.
In the first two periods there was a relatively smooth decline in the number of
women without a birth starting at the time of marriage, but in the last two
periods, there are now few women who exit early after marriage and
instead there is a substantial dip between 9 and 12 months after
marriage.

It is likely that the compressed spacing, beginning at nine months
after marriage in the later periods, is associated with better health
and higher age at marriage.
The descriptive statistics above show, for example, that women without 
education have seen an increase in the average age at marriage from below 
16 to 18.5 from the first period to the last while women with the most 
education increased from 19.5 to 20.4.

@@

For women with no education, Figure \ref{fig:pps_low} shows that the median results
above are consistent with the distribution of spacing within a spell with spacing 
after only girls consistently shorter than after only boys.
There are slight variations in how strong this effect is across the three spells.
On the one hand, for the second spell the difference in spacing between
when the first child is a girl versus when the first child is a boy as
first-born remains almost the same across the four decades.
On the other hand, for the third spell the differences between two girls and 
the other sex compositions become more pronounced over time, and especially
the difference between two girls and two boys have become substantially larger.
The fourth spell results are based on relatively few observations but do
indicate that women with either only girls or two girls and one boy have
shorter spacing than women with only boys or one girl and two boys.
Although the evidence is not substantial, it does appear that there is a slight
preference for two boys in the sense that spacing becomes longer once the couple
has at least two boys than for the other sex compositions.

@@

The distribution of spacing, shown in Figure \ref{fig:pps_high}, for the second spell 
conditional survival curve for urban women with eight or more years of education 
is almost indistinguishable from that of rural women with no education 
for the 1972--1984 period.
The small difference points to these two very different groups of women behaving similarly
in response to son preference in a situation where there was little access to
prenatal sex determination or little incentive to use it early even for those 
who might have had access.

Whereas the no education group show almost no change across the four decades,
the high education women show a rightward shift if the spacing pattern when
the first child is a girl.
The result is that the spacing patterns are close to identical after first-born 
girls and after first-born boys.
Given the elevated predicted percentage boys in Table \ref{tab:median_sex_ratio_high},
this change is consistent with the use of sex selection for women with a girl 
as their first child.

The hypothesized correlation between use of sex selection and reversal of 
the standard spacing pattern show up clearly in the 2005--2016 period for the 
third spell with spacing to the third child consistently longer with two girls 
than with either two girls or one boy and one girl.
The rightward movement for only girls from the, mostly, standard pattern in 
the 1972--1984 period to the new pattern in the 2005--2016 period is substantial.
There are also indications that the spacing distribution has moved rightward
for women with one boy and one girl relatively to women with two boys, which
provides some evidence for the use of sex selection for this group.
The caveat is that the 1972--1985 period also had a biased sex ratio, which
points to uncorrected recall error for this group.
